---
title: "2-QC"
author: "Siyuan Ma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    df_print: paged
---
# Overview
- Sensitivity check on mapping otus at a particular taxonamy rank.
- Aggregate studies into one phyloseq object.
- Library size distribution.
- Examine taxa distribution across studies (alpha diveristy).
- Uniquely present taxa.
- Uniquely absent taxa.
- Filtering.
```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```
```{r setup2, echo=FALSE}
rm(list = ls())
for(i.function in list.files("functions/", full.names = TRUE)) {
  source(i.function)
}
setup(getwd())
dir.output <- "results/2-QC/"
dir.create(dir.output, recursive = TRUE, showWarnings = FALSE)
```

```{r load datasets}
print(studies)
l_otus <- l_species <- l_genera <- list()
for(study in studies) {
  load(paste0("data/phyloseq/",
              study,
              "/otus.RData"))
  load(paste0("data/phyloseq/",
              study,
              "/species.RData"))
  load(paste0("data/phyloseq/",
              study,
              "/genera.RData"))
  l_otus[[study]] <- phylo
  l_species[[study]] <- phylo_species
  l_genera[[study]] <- phylo_genera
}
```
# Grouping otus into genera seems more appropriate than species
```{r sensitivity taxonamy}
tb_long <- l_otus %>% 
  purrr::map_dfr(
    function(phylo_otus) {
      tb_ra_otus <- phylo_otus %>% 
        transform_sample_counts(tss) %>% 
        phyloseq_to_tb()
      tb_otu_taxonamy <- phylo_otus %>% 
        tax_table2() %>% 
        tibble::as_tibble(rownames = "feature") %>% 
        dplyr::mutate(classification = dplyr::case_when(
          Rank7 != "s__" ~ "fully classified",
          Rank1 == "k__" ~ "unclassified at kingdom",
          Rank2 == "p__" ~ "unclassified at phylumn",
          Rank3 == "c__" ~ "unclassified at class",
          Rank4 == "o__" ~ "unclassified at order",
          Rank5 == "f__" ~ "unclassified at family",
          Rank6 == "g__" ~ "unclassified at genus",
          Rank7 == "s__" ~ "unclassified at species"
        ) %>% 
          factor(levels= c("fully classified",
                           "unclassified at species",
                           "unclassified at genus",
                           "unclassified at family",
                           "unclassified at order",
                           "unclassified at class",
                           "unclassified at phylumn",
                           "unclassified at kingdom")))
      
      tb_ra_otus %>% 
        dplyr::left_join(tb_otu_taxonamy,
                         by = c("feature", "Rank1", "Rank2", "Rank3", 
                                "Rank4", "Rank5", "Rank6", "Rank7"))
    }
  )
p <- tb_long %>% 
  dplyr::group_by(dataset_name, feature, classification) %>% dplyr::filter(!duplicated(feature)) %>% dplyr::summarise(n = n()) %>%
  ggplot(aes(x = dataset_name, y = n, fill = classification)) +
  geom_bar(stat = "identity") +
  rotate_xaxis(30) +
  ggtitle("OTUs are mostly classified at order, family, genus, and species level")
print(p) %>% 
  ggsave(paste0(dir.output, "fig_OTUClassification.pdf"),
         .,
         width = 8,
         height = 5)

p <- tb_long %>% 
  dplyr::group_by(dataset_name, rownames, classification) %>% dplyr::summarise(abundance = sum(abundance)) %>% dplyr::ungroup() %>% 
  dplyr::group_by(dataset_name, rownames) %>% dplyr::mutate(abundance_fullyClassified = 
                  sum(abundance[classification %in% c("fully classified",
                                                      "unclassified at species")])) %>% dplyr::ungroup() %>% 
  dplyr::arrange(dataset_name, abundance_fullyClassified) %>% dplyr::mutate(rownames = factor(rownames, levels = unique(rownames))) %>% 
  ggplot(aes(x = rownames, y = abundance, fill = classification)) +
  geom_bar(stat = "identity") +
  facet_grid(.~dataset_name, scales = "free_x", space = "free_x") +
  no_label_xaxis() +
  ggtitle("OTUs are most abundantly classified at family, genus, and species level.")
print(p) %>% 
  ggsave(paste0(dir.output, "fig_speciesClassificationAbd.pdf"),
         .,
         width = 20,
         height = 6)

tb_long_speciesUnclassified <- tb_long %>% 
  dplyr::filter(classification == "unclassified at species") %>% 
  dplyr::group_by(dataset_name, rownames, Rank6, Rank7) %>% dplyr::summarise(abundance = sum(abundance)) %>% dplyr::ungroup() %>% 
  dplyr::group_by(Rank6, Rank7) %>% dplyr::mutate(mean_abundance = mean(abundance)) %>% dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = ifelse(mean_abundance >= sort(unique(mean_abundance), decreasing = TRUE)[10],
                                  paste(Rank6, Rank7, sep = "|"), "others")) %>% 
  dplyr::arrange(dplyr::desc(mean_abundance)) %>% dplyr::mutate(taxonamy = factor(taxonamy, levels = unique(taxonamy))) %>%
  dplyr::group_by(dataset_name, rownames) %>% dplyr::mutate(most_abundant_abundance = abundance[taxonamy == levels(taxonamy)[1]]) %>% dplyr::ungroup() %>% 
  dplyr::arrange(dataset_name, most_abundant_abundance) %>% dplyr::mutate(rownames = factor(rownames, levels = unique(rownames)))
p1 <- tb_long_speciesUnclassified %>% 
  ggplot(aes(x = rownames, y = abundance, fill = taxonamy)) +
  geom_bar(stat = "identity") +
  facet_grid(.~dataset_name, scales = "free_x", space = "free_x") +
  no_label_xaxis() +
  ggtitle("Abundance of OTUs unclassified at species level")
p2 <- tb_long %>% 
  dplyr::filter(classification == "unclassified at species") %>% 
  dplyr::group_by(dataset_name) %>% 
  dplyr::filter(!duplicated(feature)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = paste(Rank6, Rank7, sep = "|")) %>% 
  dplyr::filter(taxonamy %in% levels(tb_long_speciesUnclassified$taxonamy)) %>% 
  dplyr::mutate(taxonamy = factor(taxonamy, levels = levels(tb_long_speciesUnclassified$taxonamy))) %>% 
  dplyr::group_by(dataset_name, taxonamy) %>% 
  dplyr::summarise(n_otus = n()) %>% 
  ggplot(aes(x = dataset_name, y = n_otus, fill = taxonamy)) +
  geom_bar(stat = "identity") +
  rotate_xaxis(30)
cowplot::plot_grid(p1, p2, nrow = 1, rel_widths = c(0.8, 0.2)) %>% 
  ggsave(paste0(dir.output, "fig_speciesUnclassified.pdf"),
         .,
         width = 25,
         height = 6)

tb_long_generaUnclassified <- tb_long %>% 
  dplyr::filter(classification == "unclassified at genus") %>% 
  dplyr::group_by(dataset_name, rownames, Rank5, Rank6) %>% dplyr::summarise(abundance = sum(abundance)) %>% dplyr::ungroup() %>% 
  dplyr::group_by(Rank5, Rank6) %>% dplyr::mutate(mean_abundance = mean(abundance)) %>% dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = ifelse(mean_abundance >= sort(unique(mean_abundance), decreasing = TRUE)[10],
                                  paste(Rank5, Rank6, sep = "|"), "others")) %>% 
  dplyr::arrange(dplyr::desc(mean_abundance)) %>% dplyr::mutate(taxonamy = factor(taxonamy, levels = unique(taxonamy))) %>%
  dplyr::group_by(dataset_name, rownames) %>% dplyr::mutate(most_abundant_abundance = abundance[taxonamy == levels(taxonamy)[1]]) %>% dplyr::ungroup() %>% 
  dplyr::arrange(dataset_name, most_abundant_abundance) %>% dplyr::mutate(rownames = factor(rownames, levels = unique(rownames)))
p1 <- tb_long_generaUnclassified %>% 
  ggplot(aes(x = rownames, y = abundance, fill = taxonamy)) +
  geom_bar(stat = "identity") +
  facet_grid(.~dataset_name, scales = "free_x", space = "free_x") +
  no_label_xaxis() +
  ggtitle("Abundance of OTUs unclassified at genera level")
p2 <- tb_long %>% 
  dplyr::filter(classification == "unclassified at genus") %>% 
  dplyr::group_by(dataset_name) %>% 
  dplyr::filter(!duplicated(feature)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = paste(Rank5, Rank6, sep = "|")) %>% 
  dplyr::filter(taxonamy %in% levels(tb_long_generaUnclassified$taxonamy)) %>% 
  dplyr::mutate(taxonamy = factor(taxonamy, levels = levels(tb_long_generaUnclassified$taxonamy))) %>% 
  dplyr::group_by(dataset_name, taxonamy) %>% 
  dplyr::summarise(n_otus = n()) %>% 
  ggplot(aes(x = dataset_name, y = n_otus, fill = taxonamy)) +
  geom_bar(stat = "identity") +
  rotate_xaxis(30)
p3 <- tb_long %>% 
  dplyr::filter(classification == "unclassified at genus",
                Rank5 == "f__Enterobacteriaceae",
                Rank6 == "g__") %>% 
  dplyr::group_by(dataset_name) %>% 
  dplyr::filter(!duplicated(feature)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = paste(Rank5, Rank6, sep = "|")) %>% 
  ggplot(aes(x = dataset_name)) +
  geom_bar() +
  rotate_xaxis(30) +
  ggtitle("# unclassified f_Enterobacteriacea OTUs")
cowplot::plot_grid(p1, p2, p3, nrow = 1, rel_widths = c(4, 1, 1)) %>% 
  ggsave(paste0(dir.output, "fig_generaUnclassified.pdf"),
         .,
         width = 30,
         height = 6)
```
# Examine alpha diversity.
```{r "alpha diversity across studies"}
tb_tmp <- studies %>% 
  purrr::map_dfr(function(study) {
    tibble::tibble(
      taxonamy = c("species",
                   "genera"),
      N_taxa = c(
        phyloseq::ntaxa(l_species[[study]]),
        phyloseq::ntaxa(l_genera[[study]])),
      N_samples = c(
        phyloseq::nsamples(l_species[[study]]),
        phyloseq::nsamples(l_genera[[study]])),
      median_log10LibSize = c(
        l_species[[study]] %>% 
          phyloseq::sample_sums() %>% 
          add(1) %>% 
          log10() %>% 
          median(),
        l_genera[[study]] %>% 
          phyloseq::sample_sums() %>% 
          add(1) %>% 
          log10() %>% 
          median()),
      study = study
    )
  })
p1 <- tb_tmp %>% 
  ggplot(aes(x = N_samples,
             y = N_taxa)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = study)) +
  facet_wrap(~factor(taxonamy, levels = c("species", "genera")),
             scales = "free_y")
p2 <- tb_tmp %>% 
  ggplot(aes(x = median_log10LibSize,
             y = N_taxa)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = study)) +
  facet_wrap(~factor(taxonamy, levels = c("species", "genera")),
             scales = "free_y")
print(p1)
print(p2)
plot_grid(p1, p2, nrow = 1) %>% 
  cowplot_title("Alpha diversity across studies") %>% 
  ggsave(file = paste0(dir.output, "fig_alphaDiversity.pdf"),
         width = 15,
         heigh = 5)
```
Examine uniquely present and absent taxa.
```{r "dataset specific species"}
uniqTaxa <- unique_taxa(phylo_species, name_study = "dataset_name")
uniqTaxa$tb_presence %>% 
  dplyr::filter(!duplicated(feature)) %>% 
  dplyr::group_by(n_present) %>% 
  dplyr::summarise(N = n()) %>% 
  t() %>% 
  DT::datatable()
(uniqTaxa$tb_uniqPres %>% 
    dplyr::arrange(study, mean_abundance) %>% 
    dplyr::mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(aes(x = feature, y = mean_abundance, color = study)) +
    geom_point() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))) %>% 
  ggsave(filename = paste0(dir.output, "fig_SpeciesUniquePresent.pdf"),
         plot = .,
         width = 40,
         height = 10)
(uniqTaxa$tb_uniqPres %>% 
    dplyr::filter(mean_abundance > 1e-4) %>% 
    dplyr::arrange(study, mean_abundance) %>% 
    dplyr::mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(aes(x = feature, y = mean_abundance, color = study)) +
    geom_point() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    coord_flip()) %>% 
  ggsave(filename = paste0(dir.output, "fig_SpeciesUniquePresentZoom.pdf"),
         plot = .,
         width = 15,
         height = 10)
(uniqTaxa$tb_uniqAbs %>% 
    dplyr::arrange(study, mean_abundance) %>% 
    dplyr::mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(aes(x = feature, y = mean_abundance, color = study)) +
    geom_point() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    coord_flip()) %>% 
  ggsave(filename = paste0(dir.output, "fig_SpeciesUniqueAbsent.pdf"),
         plot = .,
         width = 15,
         height = 10)

uniqTaxa <- unique_taxa(phylo_genera, name_study = "dataset_name")
uniqTaxa$tb_presence %>% 
  dplyr::filter(!duplicated(feature)) %>% 
  dplyr::group_by(n_present) %>% 
  dplyr::summarise(N = n()) %>% 
  t() %>% 
  DT::datatable()
(uniqTaxa$tb_uniqPres %>% 
    dplyr::arrange(study, mean_abundance) %>% 
    dplyr::mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(aes(x = feature, y = mean_abundance, color = study)) +
    geom_point() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))) %>% 
  ggsave(filename = paste0(dir.output, "fig_GeneraUniquePresent.pdf"),
         plot = .,
         width = 40,
         height = 10)
(uniqTaxa$tb_uniqPres %>% 
    dplyr::filter(mean_abundance > 1e-4) %>% 
    dplyr::arrange(study, mean_abundance) %>% 
    dplyr::mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(aes(x = feature, y = mean_abundance, color = study)) +
    geom_point() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    coord_flip()) %>% 
  ggsave(filename = paste0(dir.output, "fig_GeneraUniquePresentZoom.pdf"),
         plot = .,
         width = 15,
         height = 10)
(uniqTaxa$tb_uniqAbs %>% 
    dplyr::arrange(study, mean_abundance) %>% 
    dplyr::mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(aes(x = feature, y = mean_abundance, color = study)) +
    geom_point() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    coord_flip()) %>% 
  ggsave(filename = paste0(dir.output, "fig_GeneraUniqueAbsent.pdf"),
         plot = .,
         width = 15,
         height = 10)

mat_tax_tmp <- tax_table2(l_species$Jansson_Lamendella_Crohns)
mat_tax_tmp %>% 
  tibble::as_tibble(rownames = "otu_cluster") %>% 
  dplyr::filter(Rank6 == "g__Oceanobacillus")
mat_tax_tmp <- tax_table2(l_species$MucosalIBD)
mat_tax_tmp %>% 
  tibble::as_tibble(rownames = "otu_cluster") %>% 
  dplyr::filter(Rank7 == "s__blattae")
```
```{r "Filter samples"}
tb_libSize <- phylo_species %>% sample_sums() %>% 
  tibble::tibble(rownames = phyloseq::sample_names(phylo_species),
                 lib_size = .) %>% 
  dplyr::left_join(sample_data2(phylo_species) %>% 
                     tibble::as_tibble(rownames = "rownames") %>% 
                     dplyr::select(rownames = rownames,
                                   dataset_name = dataset_name),
                   by = "rownames")
p1 <- tb_libSize %>% 
  ggplot(aes(x = dataset_name, fill = (lib_size > 3000))) +
  geom_bar(stat = "count") +
  rotate_xaxis(30) +
  ggtitle("Cutoff = 3000")
p2 <- tb_libSize %>% 
  ggplot(aes(x = dataset_name, fill = (lib_size > 1000))) +
  geom_bar(stat = "count") +
  rotate_xaxis(30) +
  ggtitle("Cutoff = 1000")
plot_grid(p1, p2, nrow = 2) %>% 
  print() %>% 
  ggsave(filename = paste0(dir.output, "fig_libSize.pdf"),
         width = 10, 
         height = 10)
```

Some sanity check on library sizes.
```{r "check lib sizes", message=FALSE}
lib_size <- studies %>% 
  purrr::map_dfr(~ tibble::tibble(
    study = .x,
    sample = l_biom %>% 
      extract2(.x) %>% 
      phyloseq::sample_names(),
    lib.size = l_biom %>% 
      extract2(.x) %>% 
      phyloseq::sample_sums()
  ))  
read_counts <- studies %>% 
  purrr::map_dfr(~ paste0(dir_processed, 
                          "processed/", 
                          .x, 
                          "/16s/all_samples_read_counts.tsv") %>% 
                   readr::read_tsv(col_types = "cddd") %>% 
                   dplyr::mutate(study = .x)) 
lib_size %>% 
  dplyr::filter(sample %>% 
                  is_in(read_counts$`# sample`) %>% 
                  not) %>% 
  readr::write_tsv(paste0(dir.output, "tb_libSizeSanityCheck1.tsv")) %>% 
  DT::datatable()
read_counts %>%
  dplyr::filter(`# sample` %>% 
                  is_in(lib_size$sample) %>% 
                  not) %>% 
  readr::write_tsv(paste0(dir.output, "tb_libSizeSanityCheck2.tsv")) %>% 
  DT::datatable() # samples that aren't present in the OTU table are because of low read counts
lib_size <- lib_size %>% 
  dplyr::left_join(read_counts,
                   c("study" = "study",
                     "sample" = "# sample"))
lib_size %>% 
  ggplot(aes(x = lib.size, y = `reads mapping to OTU with taxonomy`)) +
  geom_point() +
  facet_wrap(~study) # the two are equal
lib_size %>% 
  tidyr::gather(key = "Type of read count",
                value = "N",
                "original read count",
                "lib.size",
                `reads mapping to OTU with taxonomy`,
                `reads mapping to unclassifed OTU`,
                factor_key = TRUE) %>% 
  ggplot(aes(x = study, y = log10(N + 1), color = `Type of read count`)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggsave(filename = paste0(dir.output, "fig_libSize.pdf"),
       width = 10,
       height = 8)
```

```{r "sanity check for jansson_twins_ibd"}
phylo_pipeline <- l_species[["jansson_twins_ibd"]]
phylo_qiita <- phyloseq::import_biom("~/Downloads/otu_table (1).biom")
phyloseq::sample_names(phylo_qiita) <- phyloseq::sample_names(phylo_qiita) %>% 
  stringr::str_replace_all(stringr::fixed("."), "_")
libSize_pipeline <- phyloseq::sample_sums(phylo_pipeline)
libSize_qiita <- phyloseq::sample_sums(phylo_qiita)
tibble::tibble(samples = phyloseq::sample_names(phylo_pipeline),
               libSize_pipeline = libSize_pipeline) %>% 
  dplyr::left_join(
    tibble::tibble(samples = phyloseq::sample_names(phylo_qiita),
                   libSize_qiita = libSize_qiita),
    by = "samples"
  ) %>% 
  ggplot(aes(x = libSize_pipeline,
             y = libSize_qiita)) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  geom_point()
tb_libSize_tmp <- tb_libSize %>% 
  dplyr::filter(dataset_name == "jansson_twins_ibd") %>% 
  dplyr::left_join(tibble::tibble(samples = phyloseq::sample_names(phylo_pipeline),
                                  libSize_pipeline = libSize_pipeline),
                   by = c("rownames" = "samples"))
tb_libSize_tmp %>% 
  ggplot(aes(x = lib_size,
             y = libSize_pipeline)) +
  geom_point()
otu1 <- otu_table2(phylo_species)[, "1070_CCD_p11_dc19"]
sum(otu1)
otu2 <- otu_table(phylo_pipeline)[, "1070_CCD_p11_dc19"]
sum(otu2)
test_taxTable <- tax_table2(phylo_pipeline)
test_species <- test_taxTable %>% apply(1, paste, collapse = "|")
table(test_species)
```
```{r "Filter taxa"}
phylo_species_filtered <- phylo_species %>% 
  prune_taxaSamples(kOverA2(k = phyloseq::nsamples(phylo_species) * 0.01,
                            A = 1e-4))
phylo_genera_filtered <- phylo_species_filtered %>% 
  phyloseq::tax_glom("Rank6")
save(phylo_species_filtered, file = "data/phyloseq/species_filtered.RData")
save(phylo_genera_filtered, file = "data/phyloseq/genera_filtered.RData")
```
