---
title: "2.1-Examine read depth and per-feature distributions"
author: "Siyuan Ma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    df_print: paged
---

# Overview
- Check read depth distributions
- Sensitivity check on mapping otus at a particular taxonamy rank.
- Taxa distribution across studies (alpha diveristy).
- Uniquely present/absent taxa.
- Check some individual OTUs
- Evaluate post-filtering results
- Read depth vs. number of present feature distribution
- Per feature heterogeneity.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
```

```{r setup2, include=FALSE}
smar::sourceDir("functions/", recursive = FALSE)
setup("./")
dir_output <- "results/2.1-examine_features/"
dir.create(dir_output, recursive = TRUE, showWarnings = TRUE)
```

# Load studies
```{r load datasets}
load("data/physeq/otus_prefilter.RData")
load("data/physeq/species_prefilter.RData")
load("data/physeq/genera_prefilter.RData")
```

# Examine read depths
```{r check read depths, message=FALSE}
# check original read count vs. those mapped with taxonamy vs. actual total read count
# in the OTU table. This goes into supp fig.
source("functions/manuscript/suppFigures/suppFig_readDepth.R")
p_compareReadCounts <- 
  suppFig_readDepth(physeq_species_prefilter,
                    physeq_species_prefilter,
                    physeq_species_prefilter)$p_compareReadCounts
print(p_compareReadCounts)
# median perc of reads passed filtering and mapped to taxonomy
physeq_species_prefilter %>% 
  smar::sample_data2() %>% 
  dplyr::summarise(median_prop = median(read_depth / original_read_count))
# samples passing 1000 or 3000 read depth The percentages > 3000 are in supp table.
p1 <- smar::sample_data2(physeq_species_prefilter) %>%
  ggplot(aes(x = dataset_name, fill = (read_depth > 3000))) +
  geom_bar(stat = "count") +
  smar::rotate_xaxis(30) +
  ggtitle("Cutoff = 3000")
p2 <- smar::sample_data2(physeq_species_prefilter) %>%
  ggplot(aes(x = dataset_name, fill = (read_depth > 1000))) +
  geom_bar(stat = "count") +
  smar::rotate_xaxis(30) +
  ggtitle("Cutoff = 1000")
p <- cowplot::plot_grid(p1, p2, ncol = 1) %>%
  smar::cowplot_title(title = "Read depth filtering cut off?")
print(p) %>%
  ggsave(filename = paste0(dir_output, "fig_readDepthCutoff.pdf"),
         .,
         width = 10,
         height = 8)
```

# Grouping otus into genera seems more appropriate than species

```{r sensitivity taxonamy}
tb_long <- studies %>% 
  purrr::map_dfr(
    function(study) {
      # This doesn't work!!
      # physeq_otus <- physeq_otus_prefilter %>%
      #   phyloseq::subset_samples(dataset_name == study)
      physeq_otus <- smar::sample_data2(physeq_otus_prefilter) %>%
        dplyr::filter(dataset_name %in% study) %>% 
        magrittr::extract2("sample_accession_16S") %>% 
        phyloseq::prune_samples(physeq_otus_prefilter)
      physeq_otus <- physeq_otus %>%
        phyloseq::filter_taxa(function(x) sum(x>0) >= 1, prune = TRUE)
      physeq_otus %>%
        to_relativeAbundance() %>%
        smar::phyloseq_to_tb()
    }
  ) %>% 
  dplyr::mutate(classification = dplyr::case_when(
          Rank7 != "s__" ~ "fully classified",
          Rank1 == "k__" ~ "unclassified at kingdom",
          Rank2 == "p__" ~ "unclassified at phylumn",
          Rank3 == "c__" ~ "unclassified at class",
          Rank4 == "o__" ~ "unclassified at order",
          Rank5 == "f__" ~ "unclassified at family",
          Rank6 == "g__" ~ "unclassified at genus",
          Rank7 == "s__" ~ "unclassified at species"
        ) %>%
          factor(levels= c("fully classified",
                           "unclassified at species",
                           "unclassified at genus",
                           "unclassified at family",
                           "unclassified at order",
                           "unclassified at class",
                           "unclassified at phylumn",
                           "unclassified at kingdom")))

# OTUs are mostly classified at order, family, genus, and species level
p <- tb_long %>% 
  dplyr::group_by(dataset_name, feature, classification) %>%
  dplyr::filter(!duplicated(feature)) %>% dplyr::summarise(n = dplyr::n()) %>%
  ggplot(aes(x = dataset_name, y = n, fill = classification)) +
  geom_bar(stat = "identity") +
  smar::rotate_xaxis(30) +
  ggtitle("OTUs are mostly classified at order, family, genus, and species level")
print(p) %>% 
  ggsave(paste0(dir_output, "fig_OTUClassification.pdf"),
         .,
         width = 8,
         height = 5)

# OTUs are most abundantly classified at family, genus, and species level.
p <- tb_long %>% 
  dplyr::group_by(dataset_name, rownames, classification) %>% 
  dplyr::summarise(abundance = sum(abundance)) %>% dplyr::ungroup() %>% 
  dplyr::group_by(dataset_name, rownames) %>% 
  dplyr::mutate(abundance_fullyClassified = 
                  sum(abundance[classification %in% 
                                  c("fully classified",
                                    "unclassified at species")])) %>% 
  dplyr::ungroup() %>% 
  dplyr::arrange(dataset_name, abundance_fullyClassified) %>% 
  dplyr::mutate(rownames = factor(rownames, levels = unique(rownames))) %>% 
  ggplot(aes(x = rownames, y = abundance, fill = classification)) +
  geom_bar(stat = "identity") +
  facet_grid(.~dataset_name, scales = "free_x", space = "free_x") +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank()) + 
  ggtitle("OTUs are most abundantly classified at family, genus, and species level.")
print(p) %>% 
  ggsave(paste0(dir_output, "fig_speciesClassificationAbd.pdf"),
         .,
         width = 20,
         height = 6)

# Abundance of OTUs unclassified at species level
tb_long_speciesUnclassified <- tb_long %>% 
  dplyr::filter(classification == "unclassified at species") %>% 
  dplyr::group_by(dataset_name, rownames, Rank6, Rank7) %>% 
  dplyr::summarise(abundance = sum(abundance)) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(Rank6, Rank7) %>% 
  dplyr::mutate(mean_abundance = mean(abundance)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = ifelse(mean_abundance >= 
                                    sort(unique(mean_abundance), 
                                         decreasing = TRUE)[10],
                                  paste(Rank6, Rank7, sep = "|"), 
                                  "others")) %>% 
  dplyr::arrange(dplyr::desc(mean_abundance)) %>% 
  dplyr::mutate(taxonamy = factor(taxonamy, levels = unique(taxonamy))) %>%
  dplyr::group_by(dataset_name, rownames) %>% 
  dplyr::mutate(all_abundance = sum(abundance)) %>% 
  dplyr::ungroup() %>% 
  dplyr::arrange(dataset_name, all_abundance) %>% 
  dplyr::mutate(rownames = factor(rownames, levels = unique(rownames)))
p1 <- tb_long_speciesUnclassified %>% 
  ggplot(aes(x = rownames, y = abundance, fill = taxonamy)) +
  geom_bar(stat = "identity") +
  facet_grid(.~dataset_name, scales = "free_x", space = "free_x") +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank()) + 
  ggtitle("Abundance of OTUs unclassified at species level")
p2 <- tb_long %>% 
  dplyr::filter(classification == "unclassified at species") %>% 
  dplyr::group_by(dataset_name) %>% 
  dplyr::filter(!duplicated(feature)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = paste(Rank6, Rank7, sep = "|")) %>% 
  dplyr::filter(taxonamy %in% levels(tb_long_speciesUnclassified$taxonamy)) %>%
  dplyr::mutate(taxonamy = factor(taxonamy, 
                                  levels = 
                                    levels(tb_long_speciesUnclassified$taxonamy))) %>% 
  dplyr::group_by(dataset_name, taxonamy) %>% 
  dplyr::summarise(n_otus = dplyr::n()) %>% 
  ggplot(aes(x = dataset_name, y = n_otus, fill = taxonamy)) +
  geom_bar(stat = "identity") +
  smar::rotate_xaxis(30)
cowplot::plot_grid(p1, p2, nrow = 1, rel_widths = c(0.8, 0.2)) %>% 
  ggsave(paste0(dir_output, "fig_speciesUnclassified.pdf"),
         .,
         width = 40,
         height = 10)

# Abundance of OTUs unclassified at genera level
tb_long_generaUnclassified <- tb_long %>% 
  dplyr::filter(classification == "unclassified at genus") %>% 
  dplyr::group_by(dataset_name, rownames, Rank5, Rank6) %>% 
  dplyr::summarise(abundance = sum(abundance)) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(Rank5, Rank6) %>% 
  dplyr::mutate(mean_abundance = mean(abundance)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = ifelse(mean_abundance >= 
                                    sort(unique(mean_abundance), 
                                         decreasing = TRUE)[10],
                                  paste(Rank5, Rank6, sep = "|"), 
                                  "others")) %>% 
  dplyr::arrange(dplyr::desc(mean_abundance)) %>% 
  dplyr::mutate(taxonamy = factor(taxonamy, levels = unique(taxonamy))) %>%
  dplyr::group_by(dataset_name, rownames) %>% 
  dplyr::mutate(all_abundance = sum(abundance)) %>% 
  dplyr::ungroup() %>% 
  dplyr::arrange(dataset_name, all_abundance) %>% 
  dplyr::mutate(rownames = factor(rownames, levels = unique(rownames)))
p1 <- tb_long_generaUnclassified %>% 
  ggplot(aes(x = rownames, y = abundance, fill = taxonamy)) +
  geom_bar(stat = "identity") +
  facet_grid(.~dataset_name, scales = "free_x", space = "free_x") +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank()) + 
  ggtitle("Abundance of OTUs unclassified at genera level")
p2 <- tb_long %>% 
  dplyr::filter(classification == "unclassified at genus") %>% 
  dplyr::group_by(dataset_name) %>% 
  dplyr::filter(!duplicated(feature)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = paste(Rank5, Rank6, sep = "|")) %>% 
  dplyr::filter(taxonamy %in% levels(tb_long_generaUnclassified$taxonamy)) %>% 
  dplyr::mutate(taxonamy = factor(taxonamy, 
                                  levels = 
                                    levels(tb_long_generaUnclassified$taxonamy))) %>% 
  dplyr::group_by(dataset_name, taxonamy) %>% 
  dplyr::summarise(n_otus = dplyr::n()) %>% 
  ggplot(aes(x = dataset_name, y = n_otus, fill = taxonamy)) +
  geom_bar(stat = "identity") +
  smar::rotate_xaxis(30)
p3 <- tb_long %>% 
  dplyr::filter(classification == "unclassified at genus",
                Rank5 == "f__Enterobacteriaceae",
                Rank6 == "g__") %>% 
  dplyr::group_by(dataset_name) %>% 
  dplyr::filter(!duplicated(feature)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(taxonamy = paste(Rank5, Rank6, sep = "|")) %>% 
  ggplot(aes(x = dataset_name)) +
  geom_bar() +
  smar::rotate_xaxis(30) +
  ggtitle("# unclassified f_Enterobacteriacea OTUs")
cowplot::plot_grid(p1, p2, p3, nrow = 1, rel_widths = c(4, 1, 1)) %>% 
  ggsave(paste0(dir_output, "fig_generaUnclassified.pdf"),
         .,
         width = 40,
         height = 10)
```

# Examine alpha diversity.

```{r alpha diversity across studies}
tb_nTaxaSampleReadDepth <- list(otus = physeq_otus_prefilter,
                                species = physeq_species_prefilter,
                                genera = physeq_genera_prefilter) %>% 
  purrr::imap_dfr(function(physeq, feature_level) {
    purrr::map_dfr(studies, function(study) {
      physeq <- smar::sample_data2(physeq) %>%
        dplyr::filter(dataset_name %in% study) %>% 
        magrittr::extract2("sample_accession_16S") %>% 
        phyloseq::prune_samples(physeq)
      physeq <- physeq %>%
        phyloseq::filter_taxa(function(x) any(x > 0), prune = TRUE)
      tibble::tibble(
        feature_level = feature_level,
        study = study,
        N_taxa = phyloseq::ntaxa(physeq),
        N_samples = phyloseq::nsamples(physeq),
        median_log10ReadDepth = smar::sample_data2(physeq)$read_depth %>% 
          add(1) %>% log10() %>% 
          median()
      )
    })
  })
p1 <- tb_nTaxaSampleReadDepth %>% 
  ggplot(aes(x = N_samples,
             y = N_taxa)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = study)) +
  facet_wrap(~factor(feature_level, levels = c("otus", "species", "genera")),
             scales = "free_y")
p2 <- tb_nTaxaSampleReadDepth %>% 
  ggplot(aes(x = median_log10ReadDepth,
             y = N_taxa)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = study)) +
  facet_wrap(~factor(feature_level, levels = c("otus", "species", "genera")),
             scales = "free_y")
cowplot::plot_grid(p1, p2, ncol = 1) %>% 
  smar::cowplot_title("Alpha diversity across studies") %>% 
  ggsave(file = paste0(dir_output, "fig_alphaDiversity.pdf"),
         width = 15,
         heigh = 10)
```

# Examine uniquely present and absent taxa.

```{r dataset specific species}
list(species = physeq_species_prefilter,
     genera = physeq_genera_prefilter) %>% 
  purrr::iwalk(function(physeq, feature_level) {
    tb_long <- physeq %>% 
      to_relativeAbundance() %>% 
      smar::phyloseq_to_tb()
    
    tb_presence <- tb_long %>%
      dplyr::group_by(feature, dataset_name) %>%
      dplyr::summarise(present = any(abundance > 0)) %>%
      dplyr::group_by(feature) %>%
      dplyr::mutate(n_present = sum(present),
                    n_absent = sum(!present))
    
    tb_long <- tb_long %>%
      dplyr::left_join(tb_presence,
                       by = c("feature", "dataset_name"))
    tb_uniqPres <- tb_long %>%
      dplyr::filter(n_present == 1, present) %>%
      dplyr::group_by(feature, dataset_name) %>%
      dplyr::summarise(mean_abundance = mean(abundance)) %>%
      dplyr::ungroup()
    tb_uniqAbs <- tb_long %>%
      dplyr::filter(n_absent == 1, present) %>%
      dplyr::group_by(feature) %>%
      dplyr::summarise(mean_abundance = mean(abundance)) %>%
      dplyr::ungroup() %>%
      dplyr::left_join(
        tb_presence %>%
          dplyr::filter(n_absent == 1,
                        !present),
        by = "feature"
      ) %>%
      dplyr::select(feature, dataset_name, mean_abundance)
    
    # Table summarises feature presence
    tb_presence %>%
      dplyr::filter(!duplicated(feature)) %>%
      dplyr::group_by(n_present) %>%
      dplyr::summarise(N = dplyr::n()) %>%
      t() %>%
      DT::datatable(caption = feature_level) %>%
      print()
    
    # Figure uniquely present features
    p <- tb_uniqPres %>%
      dplyr::arrange(dataset_name, mean_abundance) %>%
      dplyr::mutate(feature = factor(feature, levels = feature)) %>%
      ggplot(aes(x = feature, y = mean_abundance, color = dataset_name)) + 
      geom_point() +
      smar::rotate_xaxis(90)
    ggsave(filename = paste0(dir_output, "fig_", feature_level, "UniquePresent.pdf"),
           plot = p,
           width = 60,
           height = 20,
           limitsize = FALSE)
    p <- tb_uniqPres %>%
      dplyr::filter(mean_abundance > 1e-4) %>%
      dplyr::arrange(dataset_name, mean_abundance) %>%
      dplyr::mutate(feature = factor(feature, levels = feature)) %>%
      ggplot(aes(x = feature, y = mean_abundance, color = dataset_name)) + 
      geom_point() +
      smar::rotate_xaxis(90) + coord_flip()
    p %>% print() %>%
      ggsave(filename = paste0(dir_output, "fig_", feature_level, 
                               "UniquePresentZoom.pdf"),
             plot = .,
             width = 20,
             height = 10)
    
    # Figure uniquely absent features
    p <- tb_uniqAbs %>%
      dplyr::arrange(dataset_name, mean_abundance) %>%
      dplyr::mutate(feature = factor(feature, levels = feature)) %>%
      ggplot(aes(x = feature, y = mean_abundance, color = dataset_name)) + 
      geom_point() +
      smar::rotate_xaxis(90) + coord_flip()
    p %>% print() %>%
      ggsave(filename = paste0(dir_output, "fig_", feature_level, 
                               "UniqueAbsent.pdf"),
             plot = .,
             width = 20,
             height = 10)
  })
```

# Individual OTUS

```{r individual OTUS}
mat_tax_tmp <- smar::tax_table2(physeq_species_prefilter)
mat_tax_tmp %>% 
  tibble::as_tibble(rownames = "otu_cluster") %>% 
  dplyr::filter(Rank7 == "s__oncorhynchi")
mat_tax_tmp %>% 
  tibble::as_tibble(rownames = "otu_cluster") %>% 
  dplyr::filter(Rank7 == "s__blattae")
mat_tax_tmp %>% 
  tibble::as_tibble(rownames = "otu_cluster") %>% 
  dplyr::filter(Rank5 == "f__Enterobacteriaceae", Rank6 == "g__")
```