---
title: "4-association"
author: "Siyuan Ma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
html_document:
df_print: paged
---
# Overview
Perform association analysis on covariates.
- PERMANOVA
- Maaslin2

```{r setup, echo = FALSE}                                           
knitr::opts_knit$set(root.dir = normalizePath(".."))
```
```{r setup2, message=FALSE, warning=FALSE, echo = FALSE}
rm(list = ls())
for(i.function in list.files("functions/", full.names = TRUE)) {
  source(i.function)
}
setup(getwd())
dir_output <- "results/4-association/"
dir.create(dir_output, recursive = TRUE, showWarnings = FALSE)
```

```
# Testing for directional change on CD Behavior and UC Extent
```{r trend test}
dir_outputChunk <- paste0(dir_output, "Maaslin2/subtype_direction/")
dir.create(dir_outputChunk)
source("assets/ma_tests.R")

metadata_test <- metadata %>% 
  dplyr::mutate(race_fill = fill_na(race),
                gender_fill = fill_na(gender),
                age.cat_fill = 
                  dplyr::case_when(age < 18 ~ "child",
                                   !is.na(age) ~ "adult",
                                   TRUE ~ NA_character_) %>% fill_na,
                study_site = paste(dataset_name, 
                                   sample_type, 
                                   sep = "_"),
                subject_accession_longitudinal =
                  ifelse(dataset_name %in%
                           studies_longitudinal,
                         subject_accession,
                         NA_character_)) %>% 
  dplyr::group_by(study_site) %>% 
  dplyr::mutate(B.avail = dplyr::n_distinct(B.cat, na.rm = TRUE) == 3,
                E.avail = dplyr::n_distinct(E.cat, na.rm = TRUE) == 3) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(B23 = ifelse(B.cat %in% c("B2", "B3"),
                             1,
                             0),
                B3 = ifelse(B.cat == "B3",
                            1,
                            0),
                E23 = ifelse(E.cat %in% c("E2", "E3"),
                             1,
                             0),
                E3 = ifelse(E.cat == "E3",
                            1,
                            0))

for(feature_level in "species") {
  dir.create(paste0(dir_outputChunk, feature_level, "/"))
  load(paste0("data/phyloseq/", feature_level, "_adj.RData"))
  mat_otu_adj <- get(paste0("phylo_", feature_level, "_adj")) %>% 
    transform_sample_counts(tss) %>%  
    otu_table2
  
  l_results <- list()
  for(test in names(l_tests_subtype)) {
    print(test)
    test_param <- l_tests_subtype[[test]]
    # result <- fit_metaAnalysis(mat_otu = mat_otu_adj,
    #                            data = metadata_test %>%
    #                              dplyr::filter(!!test_param$exprs_filter),
    #                            test_variable = test_param$test_variable,
    #                            contrasts = test_param$contrasts,
    #                            batch_variable = test_param$batch_variable,
    #                            covariates = test_param$covariates,
    #                            covariates.random =
    #                              test_param$covariates.random,
    #                            moderator_variables =
    #                              test_param$moderator_variables,
    #                            n_features = 50,
    #                            normalization = "TSS",
    #                            transform = "AST",
    #                            analysis_method = "LM",
    #                            rma_method = "REML",
    #                            directory = paste0(dir_outputChunk,
    #                                               feature_level, "/",
    #                                               test, "/"),
    #                            forest.plots = TRUE,
    #                            verbose = FALSE)
    # save(result, file = paste0(dir_outputChunk, feature_level, "/",
    #                            test, "/result.RData"))
    load(paste0(dir_outputChunk, feature_level, "/",
                test, "/result.RData"))
    l_results[[test]] <- result
  }
  for(test in c("B2_vs_B1", "B3_vs_B1",
                "E2_vs_E1", "E3_vs_E1")) {
    print(test)
    load(paste0(dir_output, "Maaslin2/", feature_level, "/",
                test, "/result.RData"))
    l_results[[test]] <- result
  }
  
  plot_metaAnalysis(l_results, c("B2_vs_B1", "B3_vs_B1", "CD Behavior",
                                 "E2_vs_E1", "E3_vs_E1", "UC Extent"), 
                    "disease_subtype_direction",
                    paste0(dir_outputChunk, feature_level, "/"))
}
```
